Translation of B Implementations to the LLVM: Specification

### 2014-04-01 DOCUMENTATION/specification.tex
Adding information about the AST relating each Derived Type

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\usepackage{color}
\usepackage{minted}

@@

\section{The abstract syntax tree with support for advanced types}

\textcolor{blue}{
The supported advanced types are array and structures. The nodes Array and 
Structure can be used to declare a variable, or  declare an element, or to 
select an element in the array or the structure. 
}

@@

\begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Structure} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Structure : \B{Structure}} \\
        \hline
        \B{elements} & : & seq \B{Field} \\
	\hline
  	\multicolumn{3}{|c|}{Structure field: \B{Field}} \\
        \hline
        \B{id} & : & \B{Name} \\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record: \B{Record}} \\
        \hline
	\B{type} & : & \B{Structure}\\
	\B{values} & : & seq \B{Exp}\\
	\hline
	\multicolumn{3}{|c|}{Record field : \B{RecField}} \\
        \hline
	\B{base} & : &  \B{name} \\
	\B{element} & : & seq \B{Name} \\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Exp}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Exp} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).  Sequence and optional attributes are denoted 
      ``seq'' and ``opt'',respectively.}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

The translation presented in this paper handles the following concrete data:
\begin{description}
\item[Booleans] There is no Boolean type in LLVM, so we encode Booleans with
  one-bit integer:
$$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}.$$
\item[integers] Atelier-B offers three possible definitions for integers,
  corresponding to 16-bit, 32-bit (default) and 64-bit integers. The code
  generator has an option to specify which definition is employed, and we have:
$$\trad{INT}{\B{Type}} \isdef \llvm{i16} \mid \llvm{i32} \mid \llvm{i64}.$$
\item[enumerated sets] Again, enumerations are encoded as integer types, and the
  width of the type depends on the size of the enumeration.
$$\trad{\{e1 ... en\}}{\B{Type}} \isdef \LET w \BE \log_2 \B{n} \IN \llvm{i\PH{w}}.$$
\item[arrays] In LLVM, an array type is defined by its size and the type of its
  elements, the main issues in encoding a B array type into LLVM being to
  determine the size of the array and multi-dimensionality.

@@

  [[[ TODO ]]]
\end{description}

The current specification of the code generator does not handle deferred sets.

\section{Translation: Expressions}
\label{sec:expr}

@@

\include{AdvancedTypes}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage[colorlinks=false]{hyperref}
\usepackage{xspace,framed,here,keystroke}
\usepackage{graphicx}
\usepackage{color}
\usepackage{minted}
\usepackage{array}

@@

\section{The abstract syntax tree with support for derived types}

\textcolor{blue}{
The supported derived types are array and structures. The nodes Array and 
Structure can be used to declare a variable, or  declare an element, or to 
select an element in the array or the structure. 
}

@@

\begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Structure} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Structure : \B{Structure}} \\
        \hline
        \B{elements} & : & seq \B{Field} \\
	\hline
  	\multicolumn{3}{|c|}{Structure field: \B{Field}} \\
        \hline
        \B{id} & : & \B{Name} \\
	\B{type} & : & \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Record: \B{Record}} \\
        \hline
	\B{type} & : & \B{Structure}\\
	\B{values} & : & seq \B{Exp}\\
	\hline
	\multicolumn{3}{|c|}{Record field : \B{RecField}} \\
        \hline
	\B{base} & : &  \B{Name} \\
	\B{element} & : & seq \B{Name} \\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Exp}\\ 
	\B{arg2} & : & \B{Exp}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Exp}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Exp} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).  Sequence and optional attributes are denoted 
      ``seq'' and ``opt'',respectively.}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

The translation presented in this paper handles the following concrete data:
\begin{description}
\item[Booleans] There is no Boolean type in LLVM, so we encode Booleans with
  one-bit integer:
$$\trad{BOOL}{\B{Type}} \isdef \llvm{i1}.$$
\item[integers] Atelier-B offers three possible definitions for integers,
  corresponding to 16-bit, 32-bit (default) and 64-bit integers. The code
  generator has an option to specify which definition is employed, and we have:
$$\trad{INT}{\B{Type}} \isdef \llvm{i16} \mid \llvm{i32} \mid \llvm{i64}.$$
\item[enumerated sets] Again, enumerations are encoded as integer types, and the
  width of the type depends on the size of the enumeration.
$$\trad{\{e1 ... en\}}{\B{Type}} \isdef \LET w \BE \log_2 \B{n} \IN \llvm{i\PH{w}}.$$

\end{description}

\subsection{Derived Data Types}

The data types \textbf{structures} and \textbf{arrays} require detailed 
representation in the abstract syntax structure from the table \ref{tab:node-attr}.
The section \ref{sec:StructArray} also presents other examples and rules for these derived data types.

The following examples relate the B code and their abstract syntax representation. The \textcolor{gray}{gray elements} are not related to their abstract syntax tree to simplify the examples.

\begin{description}

@@

Declaring an structure type:
\[
  \textcolor{gray}{accountRecord \in} \underbrace{struct( \underbrace{name \in ID, balance \in \mathbb{N} }_{seq\ Field} )}_{Struct} 
\]
Initializing a record with a sequence of expressions:
\[
  \textcolor{gray}{accountRecord :=}  \underbrace{rec(\underbrace{name:xx’name, balance:(xx’balance - amt)}_{seq\ Exp})}_{Record}
\]

Setting a record field to a variable:
\[
  \textcolor{gray}{xx :=}  \underbrace{accountRecord'balance}_{RecField} 
\]

\item[arrays] In LLVM, an array type is defined by its size and the type of its
  elements, the main issues in encoding a B array type into LLVM being to
  determine the size of the array and multi-dimensionality.
Declaring an array type:
\[
\textcolor{gray}{arr \in} \underbrace{( \underbrace{(0..99) \times (0..99)}_{seq\ Set} \to \underbrace{0..1000}_{Set})}_{ArrayType}
\]
Initializing an array:
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{1 \mapsto  100,2 \mapsto 101}_{seq\ Map}\}}_{Array}
\]
\[
\textcolor{gray}{arr :=} \underbrace{\{\underbrace{\underbrace{0..100 \times 0..100  \times}_{seq\ Exp} \underbrace{\{0,1\}}_{Exp} }_{Product}\}}_{Array}
\]

Updating a selected element from array by a value of variable:
\[  \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii}_{seq\ Exp})}_{ArrayItem} \textcolor{gray}{:= xx}
\]

Setting a selected element from array to a variable:
\[ 
\textcolor{gray}{xx :=} \underbrace{\underbrace{arr}_{Array}(\underbrace{1+ii}_{seq\ Exp})}_{ArrayItem} \]
\end{description}

The current specification of the code generator does not handle deferred sets.
These data types and logic/math operators compose expressions that is in 
the next section.

\section{Translation Expressions}
\label{sec:expr}

@@

\include{DerivedTypes}
